---
title: "Modelling VO2 Data"
author: "Shaun Cameron"
date: "30/04/2020"
output: html_document
---

# Data Source

***

The data we will be using for this activity can be found in the **vo2-max_data.xlsx** file and contains data from 100 participants. It has been sourced from an excellent guide for multiple linear regression on [www.statistics.laerd.com](www.statistics.laerd.com) and consists of the following variables:

**1.** subject_id: unique subject identification number

**2.** age: subject age in years

**3.** weight: subject weight in kilograms

**4.** heart_rate: subject average heart rate in the last 5 minutes of a 20 minute submaximal cycling test

**5.** gender: 1 = male, 0 = female

**6.** vo2_max: VO2max in ml/kg/min recorded from a maximal treadmill test

Each row is one subjects data.

***
***

# Our Question

***

This data has been captured to develop a model to estimate VO2max (response variable) based on a persons age, weight, gender and heart rate during a 20 minute submaximal cycling test (explanatory/predictor variables). This would allow future participants to not have to complete a strenuous maximal running treadmill test (which might deter some participants) and rather, to just complete a less strenuous submaximal test.

We can use multiple linear regression to develop this model.


***
***


# Reading in the Data

***

```{r load packages, message=FALSE}

library(tidyverse) # loads tidyverse package

library(broom) # loads broom package

library(readxl) # loads readxl package


```

***

```{r read in data, results='hide'}

dat <- read_excel("data/vo2-max_data.xlsx") # reads in data file, and saves to object dat

```

***

```{r checking str}

str(dat) # checks structure of dat

```

***

```{r check head}

head(dat) # shows first 6 rows of dat

```

***

```{r check tail}

tail(dat) # shows last 6 rows of data

```

***
***

# Data Transformation

***

As gender is a categorical vector, its class must be changed from 'numeric' to 'factor'.

***

```{r change vector}

dat$gender <- as.factor(dat$gender) # changes gender column in data set from numerical vector to factor vector

str(dat) # shows new structure of dat

```